\vskip-0.6in Trade Agreements in the Shadow of Lobbying\protect\footnote{Formerly circulated under the title ``Trade Agreements, Lobbying and Separation of Powers''}

### 2016-06-23 draft.tex
remove G&G appendix and ref to isomorphism in sec:gh
change 'producer surplus' to 'profits' throughout text and appendix

$P_X^W$ and $P_Y^W$ are decreasing in $\tau$ and $\tau^*$ respectively, while $P_X$ and $P_Y^*$ are increasing in the respective domestic tariff. This gives rise to a terms-of-trade externality. Profits and producer surplus (identical in this model) in a sector are increasing in the price of its good and also in the domestic tariff. This fact, combined with the assumptions on specific factor ownership, motivates political activity by import-competing lobbies.

@@

\subsection{Preferences}
\label{sec:pref}
With the structure of politics and the economy symmetric and the latter fully separable, I focus on the home country's and the $X$-sector. The details are analogous for $Y$ and foreign. The home legislature's welfare function is
\begin{equation}
  W_{\mathit{ML}} = \mathit{CS}_X(\tau) + \ga(e,\ve) \cdot \mathit{PS}_X(\tau) + \mathit{CS}_Y(\tau^*) + \mathit{PS}_Y(\tau^*) + \mathit{TR}(\tau)
  \label{eq:ml}
\end{equation}
where $\mathit{CS}$ is consumer surplus, $\mathit{PS}$ is producer surplus, $\ga(e,\ta)$ is the weight placed on producer surplus in the import-competing industry, and $\mathit{TR}$ is tariff revenue.\footnote{Labor income $l$ could also be included in both median legislator and executive welfare. I omit it because its inclusion alters none of the results and only serves to complicate the exposition.} I model the decisions of the legislature as being taken by a median legislator with the weight the median legislator places on import-competing industry profits affected by the level of lobbying effort $e$ and a random variable $\ta$. I make the following assumptions on $\ga(e,\ta)$:

@@

Given its expectations and the legislature's preferences, the home lobby chooses its lobbying effort ($e_b$ to influence the break decision and $e_n$ to influence the trade war tariff) to maximize the welfare function:
\begin{equation}
  \expect \left[U_L \right] = \Pr\left[ \text{TradeWar} \right] \left[ \pi(\tau^{\mathit{n}}) - e_n \right] + \Pr\left[ \text{TradeAgreement} \right] \pi(\tau^a) - e_b
  \label{eq:lv}
\end{equation}
where $\pi(\cdot)$ is the current-period profit.

@@

The lower world price for home's export good has a larger negative effect on producer surplus than the positive effect on consumer surplus. The net effect of an increase in foreign trade-agreement tariffs on home legislative welfare is negative, leading to an increased probability that the trade agreement will be broken.

@@

I show in the \hyperlink{pfs}{Appendix} that choosing $\ga(e) = 1 + \frac{e}{a\pi_X(\tau)}$ aligns the political objective function used in the current model with the PFS framework. Although using the $\ga$-weighted government welfare function is a reduced-form approach in the sense that legislative dynamics are not fully modeled, a benefit is that it does not require returns to lobbying effort to be constant. The more flexible form of the weighting function also removes the equilibrium indeterminacies and inability to speak to distributional questions that are implied by the linear form and its transferable utility (Dixit, Grossman and Helpman 1997).

@@

\noindent The price of good $Y$ is decreasing in $\tau^{*a}$, so consumer surplus is increasing and producer surplus is decreasing in $\tau^{*a}$. Because $Y$ is being exported, the decrease in producer surplus is larger than the increase in consumer surplus, making the numerator negative. Since producer surplus in the denominator is increasing in $\tau$, the expression in Equation~\ref{appex:astar} is negative for all $\tau^{*a}$.  $\hfill\blacksquare$

@@

Because total social welfare is maximized at $\tau^a = \tau^{*a} = 0$, $W(\btn)-W(\bta)$ is always negative, whereas producer surplus is increasing in $\tau^a$, so the second summand is positive everywhere. With a positive denominator, we thus have that the derivative is positive on $[0,\tau^O]$.

@@

Thus the first summand in the numerator is zero at $\bta = \bm{0}$ and increasingly negative as $\bta$ increases. The second summand is positive everywhere because social welfare, $W$, is lowest at $\btn$ and producer surplus is increasing everywhere. Thus the numerator is positive at $\bm{0}$ and at least for very small $\bta$

@@

\vskip.4in
\noindent \textbf{\hypertarget{pfs}{Connection to Grossman and Helpman's `Protection for Sale'}} \\
To see the relationship between the two forms, write the PFS welfare function (replacing $C$ with $e$) as
\[
  e + aW = e + a \left[ \mathit{CS}_X(\tau) + \mathit{PS}_X(\tau) + \mathit{CS}_Y(\tau^*) + \mathit{PS}_Y(\tau^*) + \mathit{TR}(\tau) \right]
\]
The relationship between $e$ and the weight the legislature places on profits is endogenous; because the PFS government welfare function has such an elegant and completely specified form, it is possible to solve for equilibrium behavior to retrieve this relationship directly.

@@

In the model of this paper, this indifference condition, rearranged to solve for $\ga(e)$, is
		    \[
		      \ga(e) = \frac{\mathit{CS}_X(0) + \pi_X(0) + \mathit{CS}_Y(0) + \pi_Y(0) + \mathit{TR}(0) - \mathit{CS}_X(\tau) - \mathit{CS}_Y(\tau^*) - \pi_Y(\tau^*) - \mathit{TR}(\tau)}{\pi_X(\tau)}
		    \]

>>>>

$P_X^W$ and $P_Y^W$ are decreasing in $\tau$ and $\tau^*$ respectively, while $P_X$ and $P_Y^*$ are increasing in the respective domestic tariff. This gives rise to a terms-of-trade externality. Profits in a sector are increasing in the price of its good and also in the domestic tariff. This fact, combined with the assumptions on specific factor ownership, motivates political activity by import-competing lobbies.

@@

\subsection{Preferences}
\label{sec:pref}
With the structure of politics and the economy symmetric and the latter fully separable, I focus on the home country's and the $X$-sector. The details are analogous for $Y$ and foreign. The home legislature's welfare function is
\begin{equation}
  W_{\mathit{ML}} = \mathit{CS}_X(\tau) + \ga(e,\ve) \cdot \pi_X(\tau) + \mathit{CS}_Y(\tau^*) + \pi_Y(\tau^*) + \mathit{TR}(\tau)
  \label{eq:ml}
\end{equation}
where $\mathit{CS}$ is consumer surplus, $\pi$ are \textbf{profits}, $\ga(e,\ta)$ is the weight placed on \textbf{profits} in the import-competing industry, and $\mathit{TR}$ is tariff revenue.\footnote{Labor income $l$ could also be included in both median legislator and executive welfare. I omit it because its inclusion alters none of the results and only serves to complicate the exposition.} I model the decisions of the legislature as being taken by a median legislator with the weight the median legislator places on import-competing industry profits affected by the level of lobbying effort $e$ and a random variable $\ta$. I make the following assumptions on $\ga(e,\ta)$:

@@

Given its expectations and the legislature's preferences, the home lobby chooses its lobbying effort ($e_b$ to influence the break decision and $e_n$ to influence the trade war tariff) to maximize the welfare function:
\begin{equation}
  \expect \left[U_L \right] = \Pr\left[ \text{TradeWar} \right] \left[ \pi(\tau^{\mathit{n}}) - e_n \right] + \Pr\left[ \text{TradeAgreement} \right] \pi(\tau^a) - e_b
  \label{eq:lv}
\end{equation}
where $\pi(\cdot)$ is the current-period profit \textbf{and the subscript $X$ has been dropped for notational convenience}.

@@

The lower world price for home's export good has a larger negative effect on \textbf{profits} than the positive effect on consumer surplus. The net effect of an increase in foreign trade-agreement tariffs on home legislative welfare is negative, leading to an increased probability that the trade agreement will be broken.

@@

Although using the $\ga$-weighted government welfare function is a reduced-form approach in the sense that legislative dynamics are not fully modeled, a benefit is that it does not require returns to lobbying effort to be constant. The more flexible form of the weighting function also removes the equilibrium indeterminacies and inability to speak to distributional questions that are implied by the linear form and its transferable utility (Dixit, Grossman and Helpman 1997).

@@

\noindent The price of good $Y$ is decreasing in $\tau^{*a}$, so consumer surplus is increasing and \textbf{profits} are decreasing in $\tau^{*a}$. Because $Y$ is being exported, the decrease in \textbf{profits} is larger than the increase in consumer surplus, making the numerator negative. Since \textbf{profits} in the denominator are increasing in $\tau$, the expression in Equation~\ref{appex:astar} is negative for all $\tau^{*a}$.  $\hfill\blacksquare$

@@

Because total social welfare is maximized at $\tau^a = \tau^{*a} = 0$, $W(\btn)-W(\bta)$ is always negative, whereas \textbf{profits} are increasing in $\tau^a$, so the second summand is positive everywhere. With a positive denominator, we thus have that the derivative is positive on $[0,\tau^O]$.

@@

Thus the first summand in the numerator is zero at $\bta = \bm{0}$ and increasingly negative as $\bta$ increases. The second summand is positive everywhere because social welfare, $W$, is lowest at $\btn$ and \textbf{profits} are increasing everywhere. Thus the numerator is positive at $\bm{0}$ and at least for very small $\bta$

@@

\begin{comment}
\vskip.4in
\noindent \textbf{\hypertarget{pfs}{Connection to Grossman and Helpman's `Protection for Sale'}} \\
To see the relationship between the two forms, write the PFS welfare function (replacing $C$ with $e$) as
\[
  e + aW = e + a \left[ \mathit{CS}_X(\tau) + \mathit{PS}_X(\tau) + \mathit{CS}_Y(\tau^*) + \mathit{PS}_Y(\tau^*) + \mathit{TR}(\tau) \right]
\]
The relationship between $e$ and the weight the legislature places on profits is endogenous; because the PFS government welfare function has such an elegant and completely specified form, it is possible to solve for equilibrium behavior to retrieve this relationship directly.

@@

In the model of this paper, this indifference condition, rearranged to solve for $\ga(e)$, is
		    \[
		      \ga(e) = \frac{\mathit{CS}_X(0) + \pi_X(0) + \mathit{CS}_Y(0) + \pi_Y(0) + \mathit{TR}(0) - \mathit{CS}_X(\tau) - \mathit{CS}_Y(\tau^*) - \pi_Y(\tau^*) - \mathit{TR}(\tau)}{\pi_X(\tau)}
		    \] 
\end{comment}
